% NeurIPS 2026 Paper Checklist
% Aligned with unified_paper.tex main-text checklist (L762-783)

\section*{NeurIPS Paper Checklist}

\begin{enumerate}

\item {\bf Claims}
\begin{itemize}
    \item[$\checkmark$] All main claims (Moral Commitment Spectrum, Nash Trap, unconditional commitment necessity) are supported by experiments within the tested PGG environments (Sections~3--5). We do not claim generality beyond the evaluated settings.
    \item[$\checkmark$] Limitations are explicitly discussed in Section~5 (``Limitations and Future Work'').
\end{itemize}

\item {\bf Theory}
\begin{itemize}
    \item[$\checkmark$] Meta-ranking reward function includes all assumptions (Definition~1, Eqs.~1--2).
    \item[$\checkmark$] Group-level ESS stated as Proposition~1 with explicit conditions (group replacement dynamics, migration bound $m < m^*$).
    \item[$\checkmark$] Proof sketch for Nash Trap gradient vanishing at $\lambda \approx 0.5$ (Appendix~A).
\end{itemize}

\item {\bf Experiments}
\begin{itemize}
    \item[$\checkmark$] Three environment classes tested: Linear PGG, Non-linear PGG with tipping points, Spatial grid PGG (5$\times$5).
    \item[$\checkmark$] Agent scales: $N = 20$, $50$, $100$.
    \item[$\checkmark$] Seeds: 5--10 per experiment, specified per table caption.
    \item[$\checkmark$] Error bars represent $\pm 1$ standard deviation across seeds where shown.
    \item[$\checkmark$] Sensitivity analysis across multiplier $M \in [1.2, 2.0]$ and recovery function parameters (12 conditions; Appendix~C).
\end{itemize}

\item {\bf Reproducibility}
\begin{itemize}
    \item[$\checkmark$] Anonymized code available at \url{https://anonymous.4open.science/r/EthicaAI/}. Code also included as supplementary ZIP.
    \item[$\checkmark$] Reproduction entry point: \texttt{python scripts/reproduce.py} runs all analysis modules.
    \item[$\checkmark$] All hyperparameters specified in Table~7 (Appendix~G) and in code constants.
    \item[$\checkmark$] Random seeds explicitly set and documented per experiment.
\end{itemize}

\item {\bf Broader Impact}
\begin{itemize}
    \item[$\checkmark$] Positive: Framework for calibrating cooperative constraints in multi-agent systems with environment-specific tuning.
    \item[$\checkmark$] Risk: The commitment function $g(\theta, R)$ encodes particular value judgments; determining \emph{whose} values to encode requires interdisciplinary collaboration.
    \item[$\checkmark$] Mitigation: Dynamic $\lambda_t$ is fully transparent and auditable; no opponent parameter access required.
\end{itemize}

\item {\bf Safeguards}
\begin{itemize}
    \item[$\checkmark$] No human subjects involved; all experiments are computational simulations. [N/A]
\end{itemize}

\item {\bf Licenses}
\begin{itemize}
    \item[$\checkmark$] Code released under MIT License.
    \item[$\checkmark$] No external datasets used; all data generated by simulation.
\end{itemize}

\item {\bf LLM Usage}
\begin{itemize}
    \item[$\checkmark$] LLMs were used solely for English proofreading and \LaTeX\ formatting assistance. All scientific content, experimental design, and analysis are the authors' original work.
\end{itemize}

\end{enumerate}
